Semantic integration of physiology phenotypes: application to the cellular phenotype ontology

### 2012-01-20 trunk/paper/drinking-sr.tex

\title{Semantic integration of physiology
  phenotypes: an application to the cellular phenotype ontology}

\author{Robert Hoehndorf$^{1,*}$, Midori
  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and\\
  Georgios V. Gkoutos$^{1}$}

@@

\bibliographystyle{unsrt}
\bibliography{lc}

>>>>

\title{Semantic integration of physiology
  phenotypes: application to the cellular phenotype ontology}

\author{Robert Hoehndorf$^{1,*}$, Midori
  A. Harris$^2$,\\Heinrich Herre$^3$, Gabriella Rustici$^4$ and
  Georgios V. Gkoutos$^{1}$}

@@

\bibliographystyle{unsrt}
\begin{thebibliography}{10}

\bibitem{Schofield2011}
Paul~N. Schofield, John~P. Sundberg, Robert Hoehndorf, and Georgios~V. Gkoutos.
\newblock New approaches to the representation and analysis of phenotype
  knowledge in human diseases and their animal models.
\newblock {\em Briefings in Functional Genomics}, 10(5):258--265, 2011.

\bibitem{Smith2007}
Barry Smith, Michael Ashburner, Cornelius Rosse, Jonathan Bard, William Bug,
  Werner Ceusters, Louis~J. Goldberg, Karen Eilbeck, Amelia Ireland,
  Christopher~J. Mungall, Neocles Leontis, Philippe~R. Serra, Alan Ruttenberg,
  Susanna~A. Sansone, Richard~H. Scheuermann, Nigam Shah, Patricia~L. Whetzel,
  and Suzanna Lewis.
\newblock The {OBO} {F}oundry: coordinated evolution of ontologies to support
  biomedical data integration.
\newblock {\em Nat Biotech}, 25(11):1251--1255, 2007.

\bibitem{Gruber1995}
Thomas~R. Gruber.
\newblock Toward principles for the design of ontologies used for knowledge
  sharing.
\newblock {\em International Journal of Human-Computer Studies}, 43(5-6), 1995.

\bibitem{Guarino1998}
Nicola Guarino.
\newblock Formal ontology and information systems.
\newblock In Nicola Guarino, editor, {\em Proceedings of the 1st International
  Conference on Formal Ontologies in Information Systems}, pages 3--15. IOS
  Press, 1998.

\bibitem{Bada2004}
Michael Bada, Robert Stevens, Carole Goble, Yolanda Gil, Michael Ashburner,
  Judith~A. Blake, Michael~J. Cherry, Midori Harris, and Suzanna Lewis.
\newblock A short study on the success of the gene ontology.
\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
  Web}, 1(2):235--240, February 2004.

\bibitem{go2010}
{Gene Ontology Consortium}.
\newblock The gene ontology in 2010: extensions and refinements.
\newblock {\em Nucleic acids research}, 38(Database issue):D331--335, January
  2010.

\bibitem{goble}
C.~Goble and R.~Stevens.
\newblock State of the nation in data integration for bioinformatics.
\newblock {\em Journal of Biomedical Informatics}, 41(5):687--693, 10 2008.

\bibitem{Berners-Lee2001}
T.~Berners-Lee, J.~Hendler, O.~Lassila, et~al.
\newblock {The Semantic Web}.
\newblock {\em Scientific American}, 284(5):28--37, 2001.

\bibitem{Grau2008}
B.~Grau, I.~Horrocks, B.~Motik, B.~Parsia, P.~Patelschneider, and U.~Sattler.
\newblock {OWL} 2: The next step for {OWL}.
\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
  Web}, 6(4):309--322, November 2008.

\bibitem{Hoehndorf2010patterns}
Robert Hoehndorf, Anika Oellrich, Michel Dumontier, Janet Kelso, Dietrich
  Rebholz-Schuhmann, and Heinrich Herre.
\newblock Relations as patterns: Bridging the gap between {OBO} and {OWL}.
\newblock {\em BMC Bioinformatics}, 11(1):441+, 2010.

\bibitem{Horrocks2007}
Ian Horrocks.
\newblock {OBO} flat file format syntax and semantics and mapping to {OWL}
  {W}eb {O}ntology {L}anguage.
\newblock Technical report, University of Manchester, March 2007.
\newblock \url{http://www.cs.man.ac.uk/~horrocks/obo/}.

\bibitem{Smith2004}
Cynthia~L. Smith, Carroll-Ann~W. Goldsmith, and Janan~T. Eppig.
\newblock The mammalian phenotype ontology as a tool for annotating, analyzing
  and comparing phenotypic information.
\newblock {\em Genome Biology}, 6(1):R7, 2004.

\bibitem{Robinson2008}
P.~N. Robinson, S.~Koehler, S.~Bauer, D.~Seelow, D.~Horn, and S.~Mundlos.
\newblock The human phenotype ontology: a tool for annotating and analyzing
  human hereditary disease.
\newblock {\em American journal of human genetics}, 83(5):610--615, 2008.

\bibitem{ypo}
Stacia~R. Engel, Rama Balakrishnan, Gail Binkley, Karen~R. Christie, Maria~C.
  Costanzo, Selina~S. Dwight, Dianna~G. Fisk, Jodi~E. Hirschman, Benjamin~C.
  Hitz, Eurie~L. Hong, Cynthia~J. Krieger, Michael~S. Livstone, Stuart~R.
  Miyasato, Robert Nash, Rose Oughtred, Julie Park, Marek~S. Skrzypek, Shuai
  Weng, Edith~D. Wong, Kara Dolinski, David Botstein, and J.~Michael Cherry.
\newblock {Saccharomyces Genome Database provides mutant phenotype data}.
\newblock {\em Nucleic Acids Research}, 38(suppl 1):D433--D436, 2010.

\bibitem{wpo}
Gary Schindelman, Jolene Fernandes, Carol Bastiani, Karen Yook, and Paul
  Sternberg.
\newblock Worm phenotype ontology: integrating phenotype data within and beyond
  the c. elegans community.
\newblock {\em BMC Bioinformatics}, 12(1):32, 2011.

\bibitem{Gkoutos2005}
Georgios~V. Gkoutos, Eain~C. Green, Ann-Marie~M. Mallon, John~M. Hancock, and
  Duncan Davidson.
\newblock {Using ontologies to describe mouse phenotypes.}
\newblock {\em Genome biology}, 6(1), 2005.

\bibitem{obml2011h1}
Georgios~V. Gkoutos and Robert Hoehndorf.
\newblock Ontology-based cross-species integration and analysis of
  saccharomyces cerevisiae phenotypes.
\newblock In {\em Proceedings of the 3rd Workshop for Ontologies in Biomedicine
  and Life sciences (OBML)}, October 2011.

\bibitem{Gkoutos2009b}
Georgios~V. Gkoutos, Chris Mungall, Sandra Dolken, Michael Ashburner, Suzanna
  Lewis, John Hancock, Paul Schofield, Sebastian Kohler, and Peter~N. Robinson.
\newblock Entity/quality-based logical definitions for the human skeletal
  phenome using {PATO}.
\newblock {\em Annual International Conference of the IEEE Engineering in
  Medicine and Biology Society.}, 1:7069--7072, 2009.

\bibitem{Mungall2010}
Christopher Mungall, Georgios Gkoutos, Cynthia Smith, Melissa Haendel, Suzanna
  Lewis, and Michael Ashburner.
\newblock Integrating phenotype ontologies across multiple species.
\newblock {\em Genome Biology}, 11(1):R2+, 2010.

\bibitem{Hoehndorf2010phene}
Robert Hoehndorf, Anika Oellrich, and Dietrich Rebholz-Schuhmann.
\newblock Interoperability between phenotype and anatomy ontologies.
\newblock {\em Bioinformatics}, 26(24):3112 -- 3118, 10 2010.

\bibitem{Mungall2010go}
Christopher~J. Mungall, Michael Bada, Tanya~Z. Berardini, Jennifer Deegan,
  Amelia Ireland, Midori~A. Harris, David~P. Hill, and Jane Lomax.
\newblock Cross-product extensions of the gene ontology.
\newblock {\em Journal of biomedical informatics}, February 2010.
\newblock in press.

\bibitem{Bada2007a}
M.~Bada and L.~Hunter.
\newblock Enrichment of obo ontologies.
\newblock {\em Journal of Biomedical Informatics}, 40(3):300--315, June 2007.

\bibitem{Ogren2004}
P.~V. Ogren, K.~B. Cohen, G.~K. Acquaah-Mensah, J.~Eberlein, and L.~Hunter.
\newblock The compositional structure of gene ontology terms.
\newblock {\em Pac Symp Biocomput}, pages 214--225, 2004.

\bibitem{Horridge2007}
Matthew Horridge, Sean Bechhofer, and Olaf Noppens.
\newblock Igniting the {OWL} 1.1 touch paper: The {OWL} {API}.
\newblock In {\em Proceedings of OWLED 2007: Third International Workshop on
  OWL Experiences and Directions}, 2007.

\bibitem{Kazakov2011}
Yevgeny Kazakov, Markus Kr{\"o}tzsch, and Franti\v{s}ek Siman\v{c}\'{i}k.
\newblock Unchain my $\mathcal{EL}$ reasoner.
\newblock In {\em Proceedings of the 23rd International Workshop on Description
  Logics (DL'10)}, CEUR Workshop Proceedings. CEUR-WS.org, 2011.

\bibitem{Lock2010}
John~G Lock and Staffan Strömblad.
\newblock Systems microscopy: an emerging strategy for the life sciences.
\newblock {\em Experimental Cell Research}, 316(8):1438--1444, 2010.

\bibitem{Fuchs2010}
Florian Fuchs, Gregoire Pau, Dominique Kranz, Oleg Sklyar, Christoph Budjan,
  Sandra Steinbrink, Thomas Horn, Angelika Pedal, Wolfgang Huber, and Michael
  Boutros.
\newblock Clustering phenotype populations by genome-wide {RNAi} and
  multiparametric imaging.
\newblock {\em Molecular Systems Biology}, 6, June 2010.

\bibitem{Neumann2010}
Beate Neumann, Thomas Walter, Jean-Karim Hériché, Jutta Bulkescher, Holger
  Erfle, Christian Conrad, Phill Rogers, Ina Poser, Michael Held, Urban Liebel,
  and et~al.
\newblock Phenotypic profiling of the human genome by time-lapse microscopy
  reveals cell division genes.
\newblock {\em Nature}, 464(7289):721--727, 2010.

\bibitem{Schmitz2010}
Michael H.~A. Schmitz, Michael Held, Veerle Janssens, James R.~A. Hutchins,
  Otto Hudecz, Elitsa Ivanova, Jozef Goris, Laura Trinkle-Mulcahy, Angus~I.
  Lamond, Ina Poser, Anthony~A. Hyman, Karl Mechtler, Jan-Michael Peters, and
  Daniel~W. Gerlich.
\newblock Live-cell imaging rnai screen identifies pp2a–b55$\alpha$ and
  importin-$\beta$1 as key mitotic exit regulators in human cells.
\newblock {\em Nature Cell Biology}, 12:886--893, 2010.

\bibitem{Conrad2011}
Christian Conrad, Annelie W\"{u}nsche, Tze Heng~H. Tan, Jutta Bulkescher, Frank
  Sieckmann, Fatima Verissimo, Arthur Edelstein, Thomas Walter, Urban Liebel,
  Rainer Pepperkok, and Jan Ellenberg.
\newblock Micropilot: automation of fluorescence microscopy-based imaging for
  systems biology.
\newblock {\em Nature methods}, 8(3):246--249, March 2011.

\bibitem{Herre2006}
Heinrich Herre, Barbara Heller, Patryk Burek, Robert Hoehndorf, Frank Loebe,
  and Hannes Michalek.
\newblock {G}eneral {F}ormal {O}ntology ({GFO}) -- {A} foundational ontology
  integrating objects and processes [{V}ersion 1.0].
\newblock Onto-Med Report~8, IMISE, University of Leipzig, Leipzig, Germany,
  2006.

\bibitem{Gruninger2010}
Atalay \"{O}zg\"{o}vde and Michael Gr\"{u}ninger.
\newblock Foundational process relations in bio-ontologies.
\newblock In {\em Proceeding of the 2010 conference on Formal Ontology in
  Information Systems}, pages 243--256, Amsterdam, The Netherlands, The
  Netherlands, 2010. IOS Press.

\bibitem{Barwise1989}
J.~Barwise.
\newblock {\em The Situation in Logic}.
\newblock CSLI, Stanford, CA, 1989.

\bibitem{Hilbert1918}
David Hilbert.
\newblock {A}xiomatisches {D}enken.
\newblock {\em {M}athematische {A}nnalen}, 78:405--415, 1918.

\bibitem{Hoehndorf2011models}
Robert Hoehndorf, Michel Dumontier, John~H. Gennari, Sarala Wimalaratne,
  Bernard de~Bono, Daniel~L. Cook, and Georgios~V. Gkoutos.
\newblock Integrating systems biology models and biomedical ontologies.
\newblock {\em BMC Systems Biology}, 5(1):124+, August 2011.

\bibitem{Hoehndorf2011incon}
Robert Hoehndorf, Michel Dumontier, Anika Oellrich, Dietrich Rebholz-Schuhmann,
  Paul~N. Schofield, and Georgios~V. Gkoutos.
\newblock Interoperability between biomedical ontologies through relation
  expansion, upper-level ontologies and automatic reasoning.
\newblock {\em PLOS ONE}, 6(7):e22006, July 2011.

\bibitem{Motik2009a}
Boris Motik, Rob Shearer, and Ian Horrocks.
\newblock {Hypertableau Reasoning for Description Logics}.
\newblock {\em Journal of Artificial Intelligence Research}, 36:165--228, 2009.

\bibitem{Sirin2004}
Evren Sirin and Bijan Parsia.
\newblock Pellet: An {OWL} {DL} reasoner.
\newblock In Volker Haarslev and Ralf M{\"{o}}ller, editors, {\em Proceedings
  of the 2004 International Workshop on Description Logics, DL2004, Whistler,
  British Columbia, Canada, Jun 6-8}, volume 104 of {\em CEUR Workshop
  Proceedings}, Aachen, Germany, 2004. CEUR-WS.org.

\bibitem{Tsarkov2006}
D.~Tsarkov and I.~Horrocks.
\newblock {FaCT++} description logic reasoner: System description.
\newblock {\em Lecture Notes in Computer Science (including subseries Lecture
  Notes in Artificial Intelligence and Lecture Notes in Bioinformatics)}, 4130
  LNAI:292--297, 2006.

\bibitem{el4}
F.~Baader, C.~Lutz, and B.~Suntisrivaraporn.
\newblock {CEL} -- a polynomial-time reasoner for life science ontologies.
\newblock In U.~Furbach and N.~Shankar, editors, {\em Proceedings of the 3rd
  International Joint Conference on Automated Reasoning ({IJCAR'06})}, volume
  4130 of {\em Lecture Notes in Artificial Intelligence}, pages 287--291.
  Springer-Verlag, 2006.

\bibitem{elvira}
Robert Hoehndorf, Michel Dumontier, Anika Oellrich, Sarala Wimalaratne,
  Dietrich Rebholz-Schuhmann, Paul Schofield, and Georgios~V. Gkoutos.
\newblock A common layer of interoperability for biomedical ontologies based on
  {OWL} {EL}.
\newblock {\em Bioinformatics}, 27(7):1001--1008, April 2011.

\bibitem{cbreasoner}
Yevgeny Kazakov.
\newblock Consequence-driven reasoning for {H}orn {SHIQ} ontologies.
\newblock In {\em Proceedings of the 21st International Conference on
  Artificial Intelligence (IJCAI 2009)}, pages 2040--2045, July 11-17 2009.

\bibitem{rnao}
Robert Hoehndorf, Colin Batchelor, Thomas Bittner, Michel Dumontier, Karen
  Eilbeck, Rob Knight, Chris~J. Mungall, Jane~S. Richardson, Jesse Stombaugh,
  Eric Westhof, Craig~L. Zirbel, and Neocles~B. Leontis.
\newblock The {RNA} ontology ({RNAO}): An ontology for integrating {RNA}
  sequence and structure data.
\newblock {\em Applied Ontology}, 6(1):53--89, April 2011.

\bibitem{Hoehndorf2009sequences}
Robert Hoehndorf, Janet Kelso, and Heinrich Herre.
\newblock The ontology of biological sequences.
\newblock {\em BMC bioinformatics}, 10(1):377+, November 2009.

\bibitem{Burek2006}
Patryk Burek.
\newblock {\em Ontology of Functions}.
\newblock PhD thesis, University of Leipzig, Institute of Informatics (IfI),
  2006.

\bibitem{Herre2010}
Heinrich Herre.
\newblock {G}eneral {F}ormal {O}ntology {(GFO)}: A foundational ontology for
  conceptual modelling.
\newblock In Roberto Poli, Michael Healy, and Achilles Kameas, editors, {\em
  Theory and Applications of Ontology: Computer Applications}, chapter~14,
  pages 297{--}345. Springer, Heidelberg, 2010.

\bibitem{Loebe2007}
Frank Loebe.
\newblock Abstract vs. social roles -- towards a general theoretical account of
  roles.
\newblock {\em Applied Ontology}, 2(2):127--158, 2007.

\bibitem{Ashburner2000b}
M.~Ashburner, C.~A. Ball, J.~A. Blake, D.~Botstein, H.~Butler, J.~M. Cherry,
  A.~P. Davis, K.~Dolinski, S.~S. Dwight, J.~T. Eppig, M.~A. Harris, D.~P.
  Hill, L.~Issel-Tarver, A.~Kasarskis, S.~Lewis, J.~C. Matese, J.~E.
  Richardson, M.~Ringwald, G.~M. Rubin, and G.~Sherlock.
\newblock Gene ontology: tool for the unification of biology. the gene ontology
  consortium.
\newblock {\em Nat Genet}, 25(1):25--29, May 2000.

\bibitem{Bard2005}
Jonathan Bard, Seung~Y. Rhee, and Michael Ashburner.
\newblock An ontology for cell types.
\newblock {\em Genome Biology}, 6(2), 2005.

\bibitem{Eilbeck2005}
K.~Eilbeck, S.~E. Lewis, C.~J. Mungall, M.~Yandell, L.~Stein, R.~Durbin, and
  M.~Ashburner.
\newblock The sequence ontology: A tool for the unification of genome
  annotations.
\newblock {\em Genome Biology}, 6(R55), 2005.

\end{thebibliography}

%\bibliography{lc}
